---
title: "COVID-19 in Canada"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

This [RStudio](https://posit.co/products/open-source/rstudio/) notebook provides insight into the impact of COVID-19 in Canada, using a data set made publicly
available by Health Canada.

## The Beginning

What was the first province to report positive cases of COVID-19?

```{r}
library(tidyverse)
data <- read_csv("https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv")
data |>
  # Filter these out as they are not provinces!
  filter(prname != "Canada" & prname != "Repatriated travellers") |>
  filter(totalcases>0) |>
  slice_head(n=1) |>
  select(prname,date,totalcases) |>
  rename_at("prname",~"Province") |>
  rename_at("date",~"Date") |>
  rename_at("totalcases",~"Total Cases")
```

According to the data set, the first province to report positive cases
of COVID-19 was **British Columbia.** A quick check of the facts, however, suggests that the first recorded case was in a man from **Toronto, Ontario** who returned from Wuhan, China. More can be read about the case [here.](https://www.cbc.ca/news/health/canada-coronavirus-timeline-1.5482310) The earliest cases in the data set date back to **02/08/2020,** whereas the Ontario case was reported to the public on **01/25/2020.** The reason for the discrepancy is unknown.

## Diagrams

### Total Cases and Deaths Over Time

We can get a sense for the scale of the pandemic in Canada by plotting the total
cases and deaths over time on a **logarithmic scale,** to emphasize the rate
of change in these variables.

```{r}
library(tidyverse)
library(ggthemes)

# Disable the display of scientific notation.
options(scipen=999)

data <- read_csv("https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv")

data |> transform(totalcases = as.numeric(totalcases)) |>
  filter(prname == "Canada") |>
  filter(totalcases >= 0 & numdeaths >= 0) |>
  drop_na(c(totalcases, numdeaths)) |>
  ggplot(mapping = aes(x = date, y = totalcases)) +
  geom_line(mapping = aes(color = "Total cases"),
            linewidth = 1) +
  geom_line(
    mapping = aes(x = date, y = numdeaths, color = "Total deaths"),
    linewidth = 1
  ) +
  scale_y_continuous(trans = "log10") +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
  labs(x = "Year", y = "Total (Log. Scale)", title = "Total COVID-19 Cases and Deaths") +
  facet_wrap(~ prname) +
  scale_colour_colorblind(name = "Legend")
```

### Total Cases by Province

In order to examine the impact upon the individual provinces, first let us plot the total case count for each province.

```{r}
library(tidyverse)
library(ggthemes)

data <- read_csv("https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv")

data |>
  # Filter these out as they are not provinces!
  filter(prname != "Canada" & prname != "Repatriated travellers") |>
  transform(totalcases = as.numeric(totalcases)) |>
  group_by(prname) |>
  reframe(cases = max(totalcases, na.rm = TRUE)) |>
  ggplot(aes(y=reorder(prname,cases),x=cases)) +
  geom_col(aes(fill="Total Cases",color="Total Cases")) +
  labs(y="Province",x="Total Cases",title="Total COVID-19 Cases by Province",fill="Legend") +
  guides(color="none") +
  scale_color_colorblind() +
  scale_fill_colorblind() +
  theme(legend.position = "none")
```

### Total Deaths by Province

We will plot the total deaths that occurred in each province in
a similar fashion.

```{r}
library(tidyverse)
library(ggthemes)

data <- read_csv("https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv")

data |>
  # Filter these out as they are not provinces!
  filter(prname != "Canada" & prname != "Repatriated travellers") |>
  transform(totalcases = as.numeric(totalcases)) |>
  group_by(prname) |>
  reframe(deaths = max(numdeaths, na.rm = TRUE)) |>
  ggplot(aes(y=reorder(prname,deaths),x=deaths)) +
  geom_col(aes(fill="Total Deaths",color="Total Deaths")) +
  labs(y="Province",x="Total Deaths",title="Total COVID-19 Deaths by Province",fill="Legend") +
  guides(color="none") +
  scale_color_colorblind() +
  scale_fill_colorblind() +
  theme(legend.position = "none")
```

### Percentage of Deaths vs. Total Cases

Finally, let us plot of the percentage of deaths that occurred among each province's infected population.

```{r}
library(tidyverse)
library(ggthemes)

data <- read_csv("https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv")

data |>
  # Filter these out as they are not provinces!
  filter(prname != "Canada" & prname != "Repatriated travellers") |>
  transform(totalcases = as.numeric(totalcases)) |>
  arrange(desc(totalcases)) |>
  distinct(prname,.keep_all=TRUE) |>
  select(prname,totalcases,numdeaths) |>
  mutate(percent_deaths = numdeaths / totalcases) |>
  ggplot(aes(x=reorder(prname,percent_deaths),y=percent_deaths,color="Deaths",fill="Deaths")) +
  geom_col() + 
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  scale_color_colorblind() +
  scale_fill_colorblind() + 
  guides(color="none") +
  theme(legend.position="none") +
  labs(title="Percentage of Deaths vs. Total Cases of COVID-19",x="Province",y="Percentage")
```

## Analysis

It is clear from the data that Ontario and Quebec were the provinces hardest hit by COVID-19, in terms of total case counts, as well as deaths due to the disease. Both provinces account for well over one million cases each. In addition, both provinces reported the highest number of deaths due to COVID-19, with Quebec reporting the most at over **20,000** fatalities. Such high numbers are perhaps to be expected, however, when the population of both provinces is taken into consideration. Ontario accounts for over [40%](https://en.wikipedia.org/wiki/Demographics_of_Ontario) of the entire Canadian population alone, while Quebec accounts for a little under [23%.](https://en.wikipedia.org/wiki/Demographics_of_Quebec)

The fourth diagram, illustrating the percentage of deaths due to COVID-19 versus the total case count, presents a different picture of the same data. It was clear that Ontario and Quebec dominated in terms of case numbers and total deaths due to the disease, but in terms of this percentage Ontario ranks **sixth.** British Columbia ranks the first, but in this we find some encouraging results. British Columbia experienced less than **2%** deaths out of an infected population of hundreds of thousands; in fact, no Canadian province experienced a percentage greater than 2%. This indicates that while COVID-19 certainly exploded across the country in terms of total case counts, the total death counts for each province remained low.