---
title: "COVID-19 in Canada"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

```{r}
  library(tidyverse)
  library(ggthemes)
  data <- read_csv("https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv",col_types=list(totalcases=col_double()),na=c("-",""))
  # Disable the display of scientific notation.
  options(scipen=999)
```

## Introduction

This [RStudio](https://posit.co/products/open-source/rstudio/) notebook provides insight into the impact of COVID-19 in Canada, using a data set made publicly
available by Health Canada.

## Exploring the Data

### The Beginning

What was the first province to report positive cases of COVID-19?

```{r}
data |>
  # Filter these out as they are not provinces!
  filter(prname != "Canada" & prname != "Repatriated travellers") |>
  filter(totalcases>0) |>
  slice_head(n=1) |>
  select(prname,date,totalcases) |>
  rename_at("prname",~"Province") |>
  rename_at("date",~"Date") |>
  rename_at("totalcases",~"Total Cases")
```

According to the data set, the first province to report positive cases
of COVID-19 was **British Columbia.** A quick check of the facts, however, suggests that the first recorded case was in a man from **Toronto, Ontario** who returned from Wuhan, China. More can be read about the case [here.](https://www.cbc.ca/news/health/canada-coronavirus-timeline-1.5482310) The earliest cases in the data set date back to **02/08/2020,** whereas the Ontario case was reported to the public on **01/25/2020.** The reason for the discrepancy is unknown.

### Total Cases by Province

In order to examine the impact upon the individual provinces, first let us plot the total case count for each province.

```{r}
data |>
  # Filter these out as they are not provinces!
  filter(prname != "Canada" & prname != "Repatriated travellers") |>
  filter(!is.na(totalcases)) |>
  group_by(prname) |>
  reframe(cases = max(totalcases, na.rm = TRUE)) |>
  ggplot(aes(y=fct_reorder(prname,cases),x=cases)) +
  geom_col(aes(fill="Total Cases",color="Total Cases")) +
  labs(y="Province",x="Total Cases",title="Ontario Experienced the Most COVID-19 Cases",fill="Legend",caption="Data from Health Canada") +
  scale_color_colorblind() +
  scale_fill_colorblind() +
  theme(legend.position = "none")
```

### Total Deaths by Province

We will plot the total deaths that occurred in each province in
a similar fashion.

```{r}
data |>
  # Filter these out as they are not provinces!
  filter(prname != "Canada" & prname != "Repatriated travellers") |>
  filter(!is.na(numdeaths)) |>
  group_by(prname) |>
  reframe(deaths = max(numdeaths, na.rm = TRUE)) |>
  ggplot(aes(y=fct_reorder(prname,deaths),x=deaths)) +
  geom_col(aes(fill="Total Deaths",color="Total Deaths")) +
  labs(y="Province",x="Total Deaths",title="Quebec Experienced the Most Deaths Due to COVID-19",fill="Legend",caption="Data from Health Canada") +
  scale_color_colorblind() +
  scale_fill_colorblind() +
  theme(legend.position = "none")
```

### Percentage of Patients who Died from COVID-19

Finally, let us plot of the percentage of deaths that occurred among each province's infected population.

```{r}
data |>
  # Filter these out as they are not provinces!
  filter(prname != "Canada" & prname != "Repatriated travellers") |>
  filter(!is.na(totalcases)) |>
  arrange(desc(totalcases)) |>
  distinct(prname,.keep_all=TRUE) |>
  select(prname,totalcases,numdeaths) |>
  mutate(percent_deaths = numdeaths / totalcases) |>
  ggplot(aes(x=fct_reorder(prname,percent_deaths),y=percent_deaths,color="Deaths",fill="Deaths")) +
  geom_col() + 
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  scale_color_colorblind() +
  scale_fill_colorblind() + 
  theme(legend.position="none") +
  labs(title="British Columbia Had the Largest Percentage of Patients who Died from COVID-19" |> str_wrap(60), x="Province",y="Percentage",caption="Data from Health Canada")
```

