---
title: "COVID-19 in Canada"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

This [RStudio](https://posit.co/products/open-source/rstudio/) notebook provides insight into the impact of COVID-19 in Canada, using a data set made publicly
available by Health Canada.

## The Beginning

What was the first province to report positive cases of COVID-19?

```{r}
library(tidyverse)
data <- read_csv("https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv")
data |>
  # Filter these out as they are not provinces!
  filter(prname != "Canada" & prname != "Repatriated travellers") |>
  filter(totalcases>0) |>
  slice_head(n=1) |>
  select(prname,date,totalcases) |>
  rename_at("prname",~"Province") |>
  rename_at("date",~"Date") |>
  rename_at("totalcases",~"Total Cases")
```

According to the data set, the first province to report positive cases
of COVID-19 was **British Columbia.** A quick check of the facts, however, suggests that the first recorded case was in a man from **Toronto, Ontario** who returned from Wuhan, China. More can be read about the case [here.](https://www.cbc.ca/news/health/canada-coronavirus-timeline-1.5482310) The earliest cases in the data set date back to **02/08/2020,** whereas the Ontario case was reported to the public on **01/25/2020.** The reason for the discrepancy is unknown.

## Diagrams

### 1. Total Cases and Deaths Over Time

We can get a sense for the scale of the pandemic in Canada by plotting the total
cases and deaths over time on a **logarithmic scale,** to emphasize the rate
of change in these variables.

```{r}
library(tidyverse)
library(ggthemes)

# Disable the display of scientific notation.
options(scipen=999)

data <- read_csv("https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv")

data |> transform(totalcases = as.numeric(totalcases)) |>
  filter(prname == "Canada") |>
  filter(totalcases >= 0 & numdeaths >= 0) |>
  drop_na(c(totalcases, numdeaths)) |>
  ggplot(mapping = aes(x = date, y = totalcases)) +
  geom_line(mapping = aes(color = "Total cases"),
            linewidth = 1) +
  geom_line(
    mapping = aes(x = date, y = numdeaths, color = "Total deaths"),
    linewidth = 1
  ) +
  scale_y_continuous(trans = "log10") +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
  labs(x = "Year", y = "Total (Log. Scale)", title = "Total COVID-19 Cases and Deaths") +
  facet_wrap(~ prname) +
  scale_colour_colorblind(name = "Legend")
```

### 2. Total Cases by Province

In order to examine the impact upon the individual provinces, first let us plot the total case count for each province.

```{r}
library(tidyverse)
library(ggthemes)

data <- read_csv("https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv")

data |>
  # Filter these out as they are not provinces!
  filter(prname != "Canada" & prname != "Repatriated travellers") |>
  transform(totalcases = as.numeric(totalcases)) |>
  group_by(prname) |>
  reframe(cases = max(totalcases, na.rm = TRUE)) |>
  ggplot(aes(y=reorder(prname,cases),x=cases)) +
  geom_col(aes(fill="Total Cases",color="Total Cases")) +
  labs(y="Province",x="Total Cases",title="Total COVID-19 Cases by Province",fill="Legend") +
  guides(color="none") +
  scale_color_colorblind() +
  scale_fill_colorblind() +
  theme(legend.position = "none")
```

### 3. Total Deaths by Province

We will plot the total deaths that occurred in each province in
a similar fashion.

```{r}
library(tidyverse)
library(ggthemes)

data <- read_csv("https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv")

data |>
  # Filter these out as they are not provinces!
  filter(prname != "Canada" & prname != "Repatriated travellers") |>
  transform(totalcases = as.numeric(totalcases)) |>
  group_by(prname) |>
  reframe(deaths = max(numdeaths, na.rm = TRUE)) |>
  ggplot(aes(y=reorder(prname,deaths),x=deaths)) +
  geom_col(aes(fill="Total Deaths",color="Total Deaths")) +
  labs(y="Province",x="Total Deaths",title="Total COVID-19 Deaths by Province",fill="Legend") +
  guides(color="none") +
  scale_color_colorblind() +
  scale_fill_colorblind() +
  theme(legend.position = "none")
```

### 4. Percentage of Deaths vs. Total Cases

Finally, let us plot of the percentage of deaths that occurred among each province's infected population.

```{r}
library(tidyverse)
library(ggthemes)

data <- read_csv("https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv")

data |>
  # Filter these out as they are not provinces!
  filter(prname != "Canada" & prname != "Repatriated travellers") |>
  transform(totalcases = as.numeric(totalcases)) |>
  arrange(desc(totalcases)) |>
  distinct(prname,.keep_all=TRUE) |>
  select(prname,totalcases,numdeaths) |>
  mutate(percent_deaths = numdeaths / totalcases) |>
  ggplot(aes(x=reorder(prname,percent_deaths),y=percent_deaths,color="Deaths",fill="Deaths")) +
  geom_col() + 
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  scale_color_colorblind() +
  scale_fill_colorblind() + 
  guides(color="none") +
  theme(legend.position="none") +
  labs(title="Percentage of Deaths vs. Total Cases of COVID-19",x="Province",y="Percentage")
```

## Analysis

The second and third plots reveal **Ontario** and **Quebec** as being heavily impacted by COVID-19, both in terms of total case counts and fatalities. Each province individually accounted for well over one million cases. In addition, both provinces reported a relatively high number of deaths, with Quebec having reported the most at over **20,000** fatalities. Such high numbers must be viewed in perspective, however. Ontario accounts for over [40%](https://en.wikipedia.org/wiki/Demographics_of_Ontario) of the entire Canadian population alone, while Quebec accounts for a little under [23%.](https://en.wikipedia.org/wiki/Demographics_of_Quebec) It may be that the progression of COVID-19 was in proportion for Ontario and Quebec to the rest of the country; their dominance in total case counts and deaths due to the disease may simply be a result of their populations and population densities.

The fourth plot, illustrating the percentage of deaths due to COVID-19 versus the total case count, presents a different picture of the same data. British Columbia ranks the first in terms of the percentage of COVID-19 patients who died from the disease. However, British Columbia experienced less than **2%** deaths out of an infected population of hundreds of thousands. In fact, **no** Canadian province experienced a percentage greater than **2%.** Though the first plot clearly demonstrates exponential growth, in both total case counts and total deaths across Canada due to the disease, the actual number of deaths remained consistently low; a small percentage of each province's infected population.