---
title: "COVID-19 in Canada"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

```{r}
  library(tidyverse)
  library(ggthemes)
  data <- read_csv("https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv",col_types=list(totalcases=col_double()),na=c("-",""))
  # Disable the display of scientific notation.
  options(scipen=999)
```

## Introduction

This [RStudio](https://posit.co/products/open-source/rstudio/) notebook provides insight into the impact of COVID-19 in Canada, using a data set made publicly
available by Health Canada.

## Exploring the Data

### The Beginning

What was the first province to report positive cases of COVID-19?

```{r}
data |>
  # Filter these out as they are not provinces!
  filter(prname != "Canada" & prname != "Repatriated travellers") |>
  filter(!is.na(totalcases)) |>
  filter(totalcases > 0) |>
  group_by(prname) |>
  slice_head(n=1) |>
  arrange(date) |>
  select(prname,date,totalcases) |>
  rename_at("prname",~"Province") |>
  rename_at("date",~"Date") |>
  rename_at("totalcases",~"Total Cases")
```

**British Columbia** and **Ontario** were both the earliest provinces to report positive cases of COVID-19.

### Total Cases by Province

```{r}
data |>
  # Filter these out as they are not provinces!
  filter(prname != "Canada" & prname != "Repatriated travellers") |>
  filter(!is.na(totalcases)) |>
  group_by(prname) |>
  arrange(desc(totalcases)) |>
  slice_head(n=1) |>
  ggplot(aes(y=fct_reorder(prname,totalcases),x=totalcases)) +
  geom_col(aes(fill="Cases",color="Cases")) +
  labs(y="Province",x="Cases",title="Total COVID-19 Cases",fill="Legend",caption="Data from Health Canada") +
  scale_color_colorblind() +
  scale_fill_colorblind() +
  theme(legend.position = "none")
```

**Ontario** experienced the largest total number of COVID-19 cases. This number must be viewed in perspective, however. Ontario is Canada's most populated province, with nearly [40%](https://en.wikipedia.org/wiki/Demographics_of_Ontario) of the country's population residing in that province alone. Similarly, Quebec, ranking second, is home to a little under [23%](https://en.wikipedia.org/wiki/Demographics_of_Quebec) of the total population of Canada.

### Total Deaths by Province

```{r}
data |>
  # Filter these out as they are not provinces!
  filter(prname != "Canada" & prname != "Repatriated travellers") |>
  filter(!is.na(numdeaths)) |>
  group_by(prname) |>
  arrange(desc(numdeaths)) |>
  slice_head(n=1) |>
  ggplot(aes(y=fct_reorder(prname,numdeaths),x=numdeaths)) +
  geom_col(aes(fill="Total Deaths",color="Total Deaths")) +
  labs(y="Province",x="Total Deaths",title="Total Deaths Due to COVID-19",fill="Legend",caption="Data from Health Canada") +
  scale_color_colorblind() +
  scale_fill_colorblind() +
  theme(legend.position = "none")
```

**Quebec** recorded the most deaths due to COVID-19.

### Percentage of Patients who Died from COVID-19

Finally, let us plot of the percentage of deaths that occurred among each province's infected population.

```{r}
data |>
  # Filter these out as they are not provinces!
  filter(prname != "Canada" & prname != "Repatriated travellers") |>
  filter(!is.na(totalcases)) |>
  group_by(prname) |>
  arrange(desc(totalcases)) |>
  slice_head(n=1) |>
  select(prname,totalcases,numdeaths) |>
  mutate(percent_deaths = numdeaths / totalcases) |>
  ggplot(aes(x=fct_reorder(prname,percent_deaths),y=percent_deaths,color="Deaths",fill="Deaths")) +
  geom_col() + 
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  scale_color_colorblind() +
  scale_fill_colorblind() + 
  theme(legend.position="none") +
  labs(title="Percentage of Patients who Died from COVID-19" |> str_wrap(60), x="Province",y="Percentage",caption="Data from Health Canada")
```

**British Columbia** reported the largest percentage of patients who died as a result of COVID-19. No Canadian province reported a percentage greater than, or equal to, **2%.** This demonstrates that across Canada, the total number of deaths due to COVID-19 remained a small percentage of each province's total case count.