---
title: "COVID-19 in Canada"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

This [RStudio](https://posit.co/products/open-source/rstudio/) notebook provides insight into the impact of COVID-19 in Canada, using a data set made publicly
available by Health Canada.

## The Beginning

Which province was the first to report cases of COVID-19?

```{r}
library(tidyverse)
data <- read_csv("https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv")
data |>
  # Filter these out as they are not provinces!
  filter(prname != "Canada" & prname != "Repatriated travellers") |>
  filter(totalcases>0) |>
  slice_head(n=1) |>
  select(prname,date,totalcases) |>
  rename_at("prname",~"Province") |>
  rename_at("date",~"Date") |>
  rename_at("totalcases",~"Total Cases")
```

According to the data set, the first province to report positive cases
of COVID-19 was **British Columbia.** A quick check of the facts, however, suggests that the first recorded case was in a man from **Toronto, Ontario** who returned from Wuhan, China. More can be read about the case [here.](https://www.cbc.ca/news/health/canada-coronavirus-timeline-1.5482310) The earliest cases in the data set date back to **02/08/2020,** whereas the Ontario case was reported to the public on **01/25/2020.**

## Total Cases and Deaths

We can get a sense for the scale of the pandemic in Canada by plotting the total
cases and deaths over time on a **logarithmic scale,** to emphasize the rate
of change in these variables.

```{r}
# Plot the total cases and deaths due to COVID-19 over time in Canada.

library(tidyverse)
library(ggthemes)

# Disable the display of scientific notation.
options(scipen=999)

data <- read_csv("https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv")

data |> transform(totalcases = as.numeric(totalcases)) |>
  filter(prname == "Canada") |>
  filter(totalcases >= 0 & numdeaths >= 0) |>
  drop_na(c(totalcases, numdeaths)) |>
  ggplot(mapping = aes(x = date, y = totalcases)) +
  geom_line(mapping = aes(color = "Total cases"),
            linewidth = 1) +
  geom_line(
    mapping = aes(x = date, y = numdeaths, color = "Total deaths"),
    linewidth = 1
  ) +
  scale_y_continuous(trans = "log10") +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
  labs(x = "Year", y = "Total (Log. Scale)", title = "Total COVID-19 Cases and Deaths") +
  facet_wrap(~ prname) +
  scale_colour_colorblind(name = "Legend")
```

## The Top Five

The five provinces which experienced the highest number of COVID-19 cases and deaths were as follows:

```{r}
# What are the top five Canadian provinces with the most recorded COVID-19
# cases?

library(tidyverse)
library(ggthemes)

data <- read_csv("https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv")

data |>
  # Filter these out as they are not provinces!
  filter(prname != "Canada" & prname != "Repatriated travellers") |>
  transform(totalcases = as.numeric(totalcases)) |>
  group_by(prname) |>
  reframe(cases = max(totalcases, na.rm = TRUE),deaths=max(numdeaths,na.rm = TRUE)) |>
  arrange(desc(cases),desc(deaths)) |>
  slice_head(n=5) |>
  ggplot(aes(x=reorder(prname,-cases),y=cases)) +
  geom_col(aes(fill="Cases",color="Cases")) +
  geom_col(aes(y=deaths,fill="Deaths",color="Deaths")) +
  labs(x="Province",y="Cases/Deaths",title="Top 5 COVID-19 Cases and Deaths by Province",fill="Legend") +
  guides(color="none") +
  scale_color_colorblind() +
  scale_fill_colorblind()
```

